---
title: "PCA"
author: "Aleksei Zverev"
date: '2023-04-17'
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
library(DECIPHER)
library(phyloseq)
library(ggplot2)
library(ggpubr)
library(dplyr)
library(vegan)
library(ggvegan)
library(ggrepel)

set.seed(5678)
setwd("~/Analysis/16s-amplicon-processing/")
```

## Preparation

Merge samples to level with unique nutrition data. Sort the ASVs, select top 1000 or less. Check metadata at the map, fix it, if nessesary

```{r}
ps <- readRDS("ps.RData")

# we do not have propper column in metadata, create it
ps@sam_data$BioGroup <- paste0(ps@sam_data$Group, ".", ps@sam_data$BioRepeat)

ps.m <- merge_samples(ps, "BioGroup", fun = sum)
ps.m@sam_data
```

Sample data is fucked, so read a new nutrition data instead of it

```{r}
agro <- read.csv("agrochem_map.csv")
rownames(agro) <- agro$SampleID
agro

sample_data(ps.m) <- sample_data(agro)
```

## Made CCA model

```{r}
veganifyOTU <- function(physeq){
  require(phyloseq)
  if(taxa_are_rows(physeq)){physeq <- t(physeq)}
  return(as(otu_table(physeq), "matrix"))
}

ps.top1k <- prune_taxa(names(sort(taxa_sums(ps.m), TRUE)[1:1000]), ps.m)
X <- veganifyOTU(ps.top1k)
  
vare.cca <- vegan::cca(X ~ TOC + pH + P + K + N.ammonium + N.nitrate, data=agro)
anova(vare.cca)

```

Model is invalid :( Ok, let's see closely - it can de tuned

```{r}
anova(vare.cca, by="terms") # just one valid predictor
vif.cca(vare.cca) # lots of multicollinear predictors
```
Drop predictors one by one, according to: 1) max VIF value, 2) insignificant in anova

```{r}
vare.cca <- vegan::cca(X ~ TOC + pH + K + N.ammonium + N.nitrate, data=agro) # drop P
anova(vare.cca)
anova(vare.cca, by="terms") # still one valid predictor
vif.cca(vare.cca) # keep dropping
```
keep dropping. Valid VIF is ~<10-12

```{r}
vare.cca <- vegan::cca(X ~ TOC + pH + K + N.nitrate, data=agro) # drop N.ammonium
anova(vare.cca)
anova(vare.cca, by="terms") # still one valid predictor
vif.cca(vare.cca) # keep dropping
```
VIF is good, model is valid with one significant predictor

## Plot CCA

Our model contain three types of data:

 * CCA coordinates for the samples
 * CCA coordinates for the ASVs
 * CCA coordinates for the agrochemical vectors.
 
Enrich the model using metadata: taxonomy for ASVs, metadata for samples

```{r}
cca.data <- fortify(vare.cca)

cca.taxa <- cca.data %>% 
  filter(Score == "species") %>%
  inner_join(ps.top1k@tax_table %>% data.frame() %>% mutate(OTU = rownames(.)), 
             by = c("Label" = "OTU"))
cca.biplot <- cca.data %>% 
  filter(Score == "biplot")
cca.sites <- cca.data %>% 
  filter(Score == "sites") %>% 
  inner_join(ps.m@sam_data %>% data.frame() %>% select(Location, SampleID) %>% distinct(),
             by = c("Label" = "SampleID"))
```

```{r}
# plot species
ggplot() +
  geom_point(data=cca.taxa %>% 
               inner_join(data.frame(OTU = names(taxa_sums(ps.top1k)),
                                     Total.abund = taxa_sums(ps.top1k)),
                          by = c("Label" = "OTU")),
             aes(x=CCA1, y=CCA2, color=Phylum, size=Total.abund), alpha=0.9) +
  geom_segment(data = cca.biplot, 
               aes(x = 0, xend = CCA1, y = 0, yend = CCA2), 
               alpha=0.8, color = "black",arrow = arrow(angle = 3)) +
  geom_text_repel(data = cca.biplot, aes(x=CCA1, y=CCA2, label= Label), size=4) +
  theme_light() +
  ggtitle("A: Species")

# plot samples
ggplot() +
  geom_point(data=cca.sites, aes(x=CCA1, y=CCA2, color=Location), size=3, alpha=0.7) +
  geom_segment(data = cca.biplot, 
               aes(x = 0, xend = CCA1, y = 0, yend = CCA2), 
               alpha=0.8, color = "black",arrow = arrow(angle = 3)) +
  geom_text_repel(data = cca.biplot, aes(x=CCA1, y=CCA2, label= Label), size=4) +
  theme_light() +
  ggtitle("B. Samples")
```

