---
title: "PCA"
author: "Aleksei Zverev"
date: '2023-12-27'

output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8) 

library(phyloseq)
library(ggplot2)
library(dplyr)
library(vegan)
library(ggvegan)

set.seed(5678)
setwd("~/Analysis/16s-amplicon-processing/")
```

## Preparation

Merge samples to level with unique nutrition data. Check the metadata at the map and fix it, if necessary

```{r}
ps <- readRDS("ps.RData")

# we do not have proper column in metadata, so create it
ps@sam_data$BioGroup <- paste0(ps@sam_data$Group, ".", ps@sam_data$BioRepeat)

ps.m <- merge_samples(ps, "BioGroup", fun = sum)
ps.m@sam_data
```

The sample data are fucked, so get new nutrition data instead of it...

```{r}
agro <- read.csv("agrochem_map.csv")
rownames(agro) <- agro$SampleID
agro

sample_data(ps.m) <- sample_data(agro)
```

## Made CCA model

For a CCA model sort the ASVs, and select top-1000 or less. 

```{r}
veganifyOTU <- function(physeq){
  require(phyloseq)
  if(taxa_are_rows(physeq)){physeq <- t(physeq)}
  return(as(otu_table(physeq), "matrix"))
}

ps.top1k <- prune_taxa(names(sort(taxa_sums(ps.m), TRUE)[1:1000]), ps.m)
X <- veganifyOTU(ps.top1k)
  
vare.cca <- vegan::cca(X ~ TOC + pH + P + K + N.ammonium + N.nitrate, data=agro)
anova(vare.cca)

```

Model is invalid :( Ok, let's see it closely - it can be tuned

```{r}
anova(vare.cca, by="terms") # just one valid predictor
vif.cca(vare.cca) # lots of multicollinear predictors
```

Drop predictors one by one, according to: 1) max VIF value, 2) insignificance in ANOVA

```{r}
vare.cca <- vegan::cca(X ~ TOC + pH + K + N.ammonium + N.nitrate, data=agro) # drop P
anova(vare.cca)
anova(vare.cca, by="terms") # still one valid predictor
vif.cca(vare.cca) # keep dropping
```

keep dropping. Valid VIF is ~<10-12

```{r}
vare.cca <- vegan::cca(X ~ TOC + pH + K + N.nitrate, data=agro) # drop N.ammonium
anova(vare.cca)
anova(vare.cca, by="terms") # still one valid predictor
vif.cca(vare.cca) # keep dropping
```

VIF is good, model is valid with one significant predictor

```{r}
vare.cca
```

## Plot CCA

Our model contain three types of data:

 * CCA coordinates for the samples (`vare.cca$CCA$u`)
 * CCA coordinates for the ASVs (`vare.cca$CCA$v`)
 * CCA coordinates for the agrochemical vectors (`vare.cca$CCA$biplot`).
 
Enrich the model using this information: taxonomy for ASVs, and metadata for samples

```{r}
ASVs.data <- vare.cca$CCA$v %>% 
               data.frame() %>% 
               mutate(ASV = rownames(.)) %>% 
               inner_join(data.frame(ASV = names(taxa_sums(ps.top1k)),
                                     Total.abund = taxa_sums(ps.top1k),
                                     ps.top1k@tax_table[,2], # Phylum
                                     ps.top1k@tax_table[,3], # Class
                                     ps.top1k@tax_table[,4],
                                     ps.top1k@tax_table[,5],
                                     ps.top1k@tax_table[,6]),
                          by = "ASV")

samples.data <- vare.cca$CCA$u %>% 
  data.frame() %>% 
  mutate(Names = rownames(.)) %>% 
  inner_join(ps.top1k@sam_data %>% 
               data.frame() %>% 
               mutate(Samples = rownames(.)), by = c("Names" = "Samples"))

```

Plot our data

```{r}
# plot ASVs
ggplot() +
  geom_point(data=ASVs.data,
             aes(x=CCA1, y=CCA2, color=Phylum, size=Total.abund), alpha=0.9) +
  geom_segment(data = vare.cca$CCA$biplot %>% data.frame(),
               aes(x = 0, xend = CCA1, y = 0, yend = CCA2),
               alpha=0.8, color = "black",arrow = arrow(angle = 3)) +
  geom_text(data = vare.cca$CCA$biplot %>%
                    data.frame() %>%
                    mutate(Label = rownames(.)),
            aes(x=CCA1, y=CCA2, label= Label,
                hjust = -0.5), size=4) +
  theme_light() +
  ggtitle("A: Species")

# plot samples
ggplot() +
  geom_point(data=samples.data, 
             aes(x=CCA1, y=CCA2, color=Location, shape=Replica), size=3, alpha=0.7) +
  geom_segment(data = vare.cca$CCA$biplot %>% data.frame(),
               aes(x = 0, xend = CCA1, y = 0, yend = CCA2),
               alpha=0.8, color = "black",arrow = arrow(angle = 3)) +
  geom_text(data = vare.cca$CCA$biplot %>%
                    data.frame() %>%
                    mutate(Label = rownames(.)),
            aes(x=CCA1, y=CCA2, label= Label,
                hjust = -0.5), size=4) +
  theme_light() +
  ggtitle("B. Samples")
```

It is reasonable to filter minor phyla

```{r}
ASVs.data %>% 
  group_by(Phylum) %>% 
  summarize(sum = sum(Total.abund)) %>% 
  arrange(desc(sum))


major.phyla <- ASVs.data %>% 
  group_by(Phylum) %>% 
  summarize(sum = sum(Total.abund)) %>% 
  arrange(desc(sum)) %>% 
  select(Phylum) %>% 
  head(10) %>% 
  as.vector()

# plot major ASVs
ggplot() +
  geom_point(data=ASVs.data %>% filter(Phylum %in% major.phyla$Phylum),
             aes(x=CCA1, y=CCA2, color=Phylum, size=Total.abund), alpha=0.9) +
  geom_segment(data = vare.cca$CCA$biplot %>% data.frame(),
               aes(x = 0, xend = CCA1, y = 0, yend = CCA2),
               alpha=0.8, color = "black",arrow = arrow(angle = 3)) +
  geom_text(data = vare.cca$CCA$biplot %>%
                    data.frame() %>%
                    mutate(Label = rownames(.)),
            aes(x=CCA1, y=CCA2, label= Label,
                hjust = -0.5), size=4) +
  theme_light() +
  theme(legend.position = "bottom") +
  ggtitle("A: Species")
```

... or plot different phyla separately (uncomment labels to indicate ASV names)

```{r}
for (i in major.phyla$Phylum) {
  
  if (i == "Proteobacteria") {
    for (j in c("Deltaproteobacteria", "Alphaproteobacteria", "Gammaproteobacteria")){

      p <- ggplot() +
      geom_point(data=ASVs.data,
                 aes(x=CCA1, y=CCA2, size=Total.abund), alpha=0.2, color="grey80") +
      geom_point(data=ASVs.data %>% filter(Class == j),
                 aes(x=CCA1, y=CCA2, color=Family, size=Total.abund), alpha=0.9) +
#### comment this to remove labels ####
      geom_text(data=ASVs.data %>% filter(Class == j),
                 aes(x=CCA1, y=CCA2, color=Family, label=paste0('#', substring(ASV, 4))),
                size=3, nudge_y=0.1) +
####                               ####
      geom_segment(data = vare.cca$CCA$biplot %>% data.frame(),
                  aes(x = 0, xend = CCA1, y = 0, yend = CCA2),
                  alpha=0.8, color = "black",arrow = arrow(angle = 3)) +
      geom_text(data = vare.cca$CCA$biplot %>%
                        data.frame() %>%
                        mutate(Label = rownames(.)),
                aes(x=CCA1, y=CCA2, label= Label,
                    hjust = -0.5), size=4) +
      theme_light() +
      ggtitle(paste(i, " - ",j))
    print(p)
    }
  } else {
    
    p <- ggplot() +
      geom_point(data=ASVs.data,
                 aes(x=CCA1, y=CCA2, size=Total.abund), alpha=0.2, color="grey80") +
      geom_point(data=ASVs.data %>% filter(Phylum == i),
                 aes(x=CCA1, y=CCA2, color=Family, size=Total.abund), alpha=0.9) +
#### comment this to remove labels ####
      geom_text(data=ASVs.data %>% filter(Phylum == i),
                 aes(x=CCA1, y=CCA2, color=Family, label=paste0('#', substring(ASV, 4))),
                size=3, nudge_y=0.1) +
####                               ####
      geom_segment(data = vare.cca$CCA$biplot %>% data.frame(),
                   aes(x = 0, xend = CCA1, y = 0, yend = CCA2),
                   alpha=0.8, color = "black",arrow = arrow(angle = 3)) +
      geom_text(data = vare.cca$CCA$biplot %>%
                        data.frame() %>%
                        mutate(Label = rownames(.)),
                aes(x=CCA1, y=CCA2, label= Label,
                    hjust = -0.5), size=4) +
      theme_light() +
      ggtitle(i)
    print(p)
    }
}
```








